---
title: "5-State Markov Model - Covid Staging Data"
author: "Steph Reynolds (Stephanie.Reynolds@ucsf.edu)"
date: "`r format(Sys.time(), '%b %d, %Y  %H:%M %p')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Background

## Project 
MINDSCAPE: Modeling of infectious network dynamics for surveillance, control and prevention enhancement

## Description
This file builds a 5-state Markov Model using the msm package to assess the trajectory of Covid-positive patients. The five states include: LowMod=1, HighMod=2, Severe=3, Dead=4, and Discharged=5. The resulting Markov Model will then be compared to the 4-State MSM.

# Load required packages
```{r Load required packages, include=F}
library(tidyverse)
library(msm)
```

# Import and preview data
```{r Import data}
raw <- read_csv("/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/data/dm_stg_clean_11.08.21.csv")

# Frequency table for stage
table(raw$stage)
```

# Transform stages to stage categories

    * Stage 4 -> 1 (Low Moderate)
    * Stage 5 -> 2 (High Moderate)
    * Stage 6-9 -> 3 (Severe)
    * Stage 10 -> 4 (Dead)
    * Discharged -> 5 (Discharged)
```{r Collapse stages to stage categories}
df <- raw %>%
    mutate(stagecat = case_when(stage == 4 ~ "1",
                                stage == 5 ~ "2",
                                stage %in% 6:9 ~ "3",
                                stage == 10 ~ "4")) %>%
    select(ID, date, stagecat)
```

# Prep dataset for Markov Model
For each 'ID', create new row to represent discharge state. Bind to dataset and arrange by 'ID' and 'date'.
```{r Prep data for Markov Model}
markov_w_5_stages <- df %>%
    group_by(ID) %>%
    summarize(date = max(date) + 1) %>%
    mutate(stagecat = NA) %>%
    rbind(df) %>%
    arrange(ID, date) %>%
    mutate(stagecat = case_when(is.na(stagecat) ~ "5",
                                TRUE ~ stagecat))

markov_w_5_stages$stagecat <- factor(markov_w_5_stages$stagecat, 
                          ordered = TRUE, 
                          levels = c("1", "2", "3", "4", "5"))
```
    
# Generate table of state transitions
Create freq table counting # of times each pair of states were observed in succession, where rows = starting states, and columns = ending states.
```{r State transition table}
statetable.msm(state = stagecat, subject = ID, data = markov_w_5_stages)
```

# Create matrix of "allowed" transitions
where 1 = allowed transitions and 0 = forbidden transitions 
```{r Transition matrix}
qmat1 <- matrix(c(1,1,1,1,1,
                  1,1,1,1,1,
                  1,1,1,1,1,
                  0,0,0,0,1,
                  0,0,0,0,1),
                nrow = 5, ncol = 5,
                byrow=TRUE,
                dimnames=list(from=1:5, to=1:5))
```

# Fit continuous time Markov multi-state model (MSM) by maximum likelihood
Note: Set obstype = 2 --> Considered an "exact transition time." Even though this is not necessarily true with our dataset since it's more discrete than continuous, we are assuming that the pt remains at the state at previous observation until the current observation. It does not allow for a change in state in between observation points.
```{r Fit Markov model and print}
# Run MSM function
m_w_5_states <- msm(stagecat ~ date, ID, data = markov_w_5_stages, qmatrix = qmat1, obstype = 2)

# Print MSM 
printold.msm(m_w_5_states)

# Save as R object
saveRDS(m_w_5_states, "/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/5-state_MM")
```

# Total LOS for patients at moderate (1) and severe (2) stage categories
```{r Total LOS}
totlos.msm(m_w_5_states, end = c("1", "2", "3"))
```

# Probability matrix and estimated transition intensity (rate) matrix
```{r Probability and transition intensity matrices}
pmatrix.msm(m_w_5_states)

qmatrix.msm(m_w_5_states)
```

# Log likelihood ratio
```{r Log likelihood ratio}
logLik.msm(m_w_5_states)
```

# Compare 4-State and 5-State models
```{r Compare both models w/ diff in LogLik and AIC}
# Load 4-state_MM
m_w_4_states <- readRDS("/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/4-state_MM")

# Calculate difference in log likelihood ratios between `m` and `m_w_5_states`
diff_logLik <- logLik.msm(m_w_5_states) - logLik.msm(m_w_4_states)

# Calculate standard AIC for both models
AIC(m_w_4_states, m_w_5_states)

# Difference in Restricted AIC (DRAIC) ---- 
    # Compare predictive ability of two Markov multi-state models (MSM)
    # NOTE: The functions below take forever to run.

# DRAIC_results <- draic.msm(msm.full = m_w_5_states, msm.coarse = m_w_4_states)
# draic_value <- DRAIC_results$draic

# Difference in Restricted AIC (DRAIC) = 0.144 indicating that the coarse model (one w/ fewer parameters) is preferred.
    # Positive value for DRAIC = coarse model is preferred, while negative value = full model is preferred.
```

