---
title: "Development of Markov Model - Covid Staging Data"
author: "Steph Reynolds (Stephanie.Reynolds@ucsf.edu)"
date: "`r format(Sys.time(), '%b %d, %Y  %H:%M %p')`"
output:
  pdf_document:
    toc: true
    number_sections: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Background

## Project

MINDSCAPE: Modeling of infectious network dynamics for surveillance,
control and prevention enhancement

## Description

This files builds a 4-state Markov Model using the `markovchain` package
to assess the trajectory of hospitalized COVID-19 patients. The four
states include: Moderate=1, Severe=2, Dead=3, and Discharged=4.

The stage transition matrix probabilities were imported from the
previous script.

# Load required packages

```{r Load required packages, include=F}
library(markovchain)
library(igraph)
library(tidyverse)

# Common issues with select being masked from dplyr by MASS
select <- dplyr::select
```

# Import stage transition probabilites vector and convert to matrix `prob_mat`

```{r Import stage transition prob vector and convert to matrix}
# Read in prob_vect
prob_vect <- readRDS("/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/data/prob_vect.rds")

# Assign names of stages to states vector
states <- c("Moderate", "Severe", "Dead", "Discharged")

# Convert prob_vect to matrix 
prob_mat <- matrix(data = prob_vect, byrow = T, 
                   nrow = 4,
                   ncol = 4,
                   dimnames = list(from = states, to = states))
prob_mat

# Save as RDS object
saveRDS(prob_mat, "/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/data/prob_mat.rds")
```

# Create Markov Chain object

```{r Create Markov Chain object}
mc <- new("markovchain", states = states, byrow = T, transitionMatrix = prob_mat, name = "Markov Model for COVID-19 Patients")
mc
```

# Plot Markov Chain

```{r Plot Markov Chain}
plot(mc, main = "Markov Model for COVID-19 Patients")
```

# Practice with `markovchain` package

## Define initial state probabilities and calculate likelihood of states after 5 days

```{r Practice defining initial state probs and calc likelihood of states after 5 days, eval = F}
init_state <- c(.5, .5, 0, 0)
after5Days <- (t(mc^5)) * init_state
after5Days
```

## Obtain transition probabilities

```{r Obtain transition probabilities}
# View states 
markovchain::states(mc)

# Obtain transition probabilities
    # Must provide initial state (t0) and subsequent state (t1)

    # Example: What is probability of transitioning FROM moderate to moderate?
    transitionProbability(mc, t0 = "Moderate", t1 = "Moderate")
    
    # Example: What is probability of transitioning FROM moderate to severe?
    transitionProbability(mc, t0 = "Moderate", t1 = "Severe")
```

## Convert Markov Chain to a dataframe to view all transition probabilities

```{r Convert mc to dataframe to view all trans probs}
as(mc, "data.frame")
```

## View summary of Markov Chain object to see closed, transient, and absorbing states

```{r View summary of Markov Chain object}
summary(mc)

# Other functions
markovchain::transientStates(mc)
markovchain::absorbingStates(mc)
```

# Plot Markov Chain using `igraph` package

```{r Plot Markov Chain using igraph package}
# Convert Markov Chain to igraph object `g`
g <- as(mc, "igraph")

# Set colors for vertices 
V(g)["Moderate"]$color <- "red"
V(g)["Severe"]$color <- "green"
V(g)["Dead"]$color <- "blue"
V(g)["Discharged"]$color <- "purple"

# Plot Markov Chain 
plot(g, vertex.color = V(g)$color, vertex.size = 25, vertex.label.cex = 1.25, vertex.label.family = "Helvetica", vertex.label.dist = 3.5, edge.width = 2, edge.arrow.size = 1, edge.label = E(g)$prob, main = "Markov Chain for Predicting the Trajectory of COVID-19 Patients")

## More advanced stuff.... 

# Identify and format the strongly connected nodes 
SCC <- clusters(g, mode="strong")
V(g)$color <- rainbow(SCC$no)[SCC$membership]

# Plot with highlighted strongly connected components or states
plot(g, mark.groups=split(1:vcount(g), SCC$membership), main="Markov Model\nwith strongly connected components")
```

# Generate sequence of states from Markov chain
```{r}
rep <- replicate(n=1000, expr=covid_sim_fun)

covid_sim_fun <- rmarkovchain(n = 5, object = mc, t0 = "Moderate", include.t0 = T)
covid_sim

CovidFittedMLE <- markovchainFit(data = covid_sim, method = "mle", name = "COVID MLE")
CovidFittedMLE$estimate
CovidFittedMLE$standardError
CovidFittedMLE$logLikelihood

CovidFittedBoot <- markovchainFit(data = covid_sim, method = "bootstrap", nboot = 1000)
CovidFittedBoot$estimate
CovidFittedBoot$standardError
CovidFittedBoot$logLikelihood

CovidFittedBootParallel <- markovchainFit(data = covid_sim, method = "bootstrap", nboot = 1000, parallel = T)
CovidFittedBootParallel$estimate
CovidFittedBootParallel$standardError
CovidFittedBootParallel$logLikelihood

```



Simulation

Simulating a random sequence from an underlying DTMC is quite easy thanks to the function rmarkovchain. The following code generates a year of weather states according to mcWeather underlying stochastic process.

weathersOfDays <- rmarkovchain(n = 365, object = mcWeather, t0 = "sunny")
weathersOfDays[1:30]
