---
title: "Fitting, Predicting, and Evaluating Markov Model - Covid Staging Data"
author: "Steph Reynolds (Stephanie.Reynolds@ucsf.edu)"
date: "`r format(Sys.time(), '%b %d, %Y  %H:%M %p')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Background

## Project

MINDSCAPE: Modeling of infectious network dynamics for surveillance, control and prevention enhancement

## Description

This file imports the Markov Model for the Covid staging data and generates fitted Markov chain for a given sequence.

-   **markovchainFit -** Function to return fitted Markov chain for a given sequence

-   **predict -** Function to calculate predictions from markovchain objects

-   **rmarkovchain -** Function to sample from markovchain objects

# Load required packages and import Markov chain

```{r Load required packages, message = F}
library(markovchain)

mc <- readRDS("/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/data/mc.rds")
mc
```

# Generate fitted Markov chain for given sequence 

```{r}
# Generate sequence of states from Markov chain
rep <- replicate(n=1000, expr=covid_sim_fun)

covid_sim_fun <- rmarkovchain(n = 5, object = mc, t0 = "Moderate", include.t0 = T)
covid_sim

CovidFittedMLE <- markovchainFit(data = covid_sim, method = "mle", name = "COVID MLE")
CovidFittedMLE$estimate
CovidFittedMLE$standardError
CovidFittedMLE$logLikelihood

CovidFittedBoot <- markovchainFit(data = covid_sim, method = "bootstrap", nboot = 1000)
CovidFittedBoot$estimate
CovidFittedBoot$standardError
CovidFittedBoot$logLikelihood

CovidFittedBootParallel <- markovchainFit(data = covid_sim, method = "bootstrap", nboot = 1000, parallel = T)
CovidFittedBootParallel$estimate
CovidFittedBootParallel$standardError
CovidFittedBootParallel$logLikelihood

```
