---
title: "Markov Model to Predict Clinical Trajectories of COVID-19 Patients"
author: "Steph Reynolds (Stephanie.Reynolds@ucsf.edu)"
date: "`r format(Sys.time(), '%b %d, %Y  %H:%M %p')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Background

## Project

MINDSCAPE: Modeling of infectious network dynamics for surveillance, control and prevention enhancement

## Description

This script builds and fits a Markov Model using the `msm` package to assess the clinical trajectory of hospitalized COVID-19 patients. The four states include:

1.  Moderate w/o history of severe (M)

2.  Moderate w/ history of severe (MS)

3.  Severe (S)

4.  Dead (D)

# Load required packages

```{r Load required packages, include=F}
library(msm)
library(slider) 
library(lubridate)
library(tidyverse)
```

# Import and preview data

```{r}
dm_stg_clean <- read_csv("/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/data/dm_stg_clean_11.08.21.csv")
glimpse(dm_stg_clean)
```

# Categorize patients into Moderate w/o hx of Severe (M), Moderate w/ hx of Severe (S), Severe (S), and Dead (D)

```{r Categorize into M, MS, S, and D}
df <- dm_stg_clean %>%
    mutate(stage_cat = case_when(stage %in% 4:5 ~ "M",
                                 stage %in% 6:9 ~ "S",
                                 stage == 10 ~ "D"),
           severe_yn = ifelse(stage_cat=="S", 1, 0)) %>% 
    group_by(ID) %>% 
    mutate(hx_severe = slider::slide_index_sum(x = severe_yn, 
                                               i = date, 
                                               before = lubridate::days(365)),
           state = case_when(stage_cat=="D" ~ "D",
                             stage_cat=="S" ~ "S",
                             stage_cat=="M" & hx_severe==0 ~ "M",
                             stage_cat=="M" & hx_severe>0 ~ "MS")) %>% 
    select(ID, date, state, age, sex, BMI, race, ethnicity)
    # Can choose to include additional covariates here
```

# Build and fit baseline model

## Create stage transition table

```{r Stage transition table}
statetbl <- statetable.msm(state, ID, data=df)
statetbl
```

## Assign matrix of allowed transitions, where 0=not allowed and 1=allowed

```{r Matrix of allowed transitions}
qmat <- rbind(c(0,0,0,0),
              c(1,1,0,1),
              c(1,0,1,1),
              c(1,0,1,1))

# Assign row and column names to qmat 
rownames(qmat) <- c("D",rownames(statetbl))
colnames(qmat) <- colnames(statetbl)

qmat
```

## Reformat `state` variable and create variable `DaysSinceEntry`

```{r Create DaysSinceEntry}
df$state <- as.numeric(as.factor(df$state))

#df$JulianDay <- as.numeric(df$date-min(df$date))

df <- df %>% 
    group_by(ID) %>% 
    mutate(DaysSinceEntry = as.numeric(date-min(date)))
```

## Specify initial values and assign to `Q.crude`

```{r Initial values}
Q.crude <- crudeinits.msm(state ~ DaysSinceEntry, ID, data=df, qmatrix=qmat)
```

## Fit baseline model and summarize

```{r Baseline model}
m0 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude)
summary.msm(m0)
```

# Build and fit covariate models

## Age

```{r Age}
df$age.cen <- df$age/100
df$age.mtert <- (df$age > quantile(df$age, 1/3, na.rm=T) & df$age<=quantile(df$age, 2/3, na.rm=T)) + 0
df$age.utert <- (df$age > quantile(df$age, 2/3, na.rm=T)) + 0


m1 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
          covariates = ~ age.cen)

m1b <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
           covariates = list("2-4" = ~ age.mtert + age.utert,
                             "3-4" = ~ age.mtert + age.utert,
                             "4-3" = ~ age.mtert + age.utert,
                             "4-1" = ~ age.mtert + age.utert))

m1a <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
           covariates = list("2-4" = ~ age.cen,
                             "3-4" = ~ age.cen,
                             "4-3" = ~ age.cen,
                             "4-1" = ~ age.cen))

summary.msm(m1a)

# Evaluate log likelihoods and compare 
pvalue.age <- 1 - pchisq((logLik(m1a) - logLik(m0)) * 2, df=3)
    # Tells us that age is important; sig P value
    # Will need data on vacc status - age liked confounded by vacc status 

# Evaluate AIC and compare
AIC(m0, m1, m1a)
```

## BMI

```{r BMI}
df$BMI.cen <- df$BMI/100

m2 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
          covariates = ~ BMI.cen)

m2a <- msm(state ~ DaysSinceEntry, subject=ID, data = df, qmatrix = Q.crude,
           covariates = list("2-4" = ~ BMI.cen,
                             "3-4" = ~ BMI.cen,
                             "4-3" = ~ BMI.cen,
                             "4-1" = ~ BMI.cen))

summary.msm(m2a)

# Evaluate AIC and compare
AIC(m0, m2, m2a)
```

## Sex

```{r Sex}
df <- df %>% 
    mutate(Male = ifelse(sex=="Male", 1, 0),
           Female = ifelse(sex=="Female", 1, 0))

m3 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude, covariates = ~ Male)

m3 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
          covariates = ~ as.factor(sex))

summary.msm(m3)
```

## Race

```{r Race}
df <- df %>% 
    mutate(Black = ifelse(race=="Black or African American", 1, 0),
           Asian = ifelse(race=="Asian", 1, 0),
           White = ifelse(race=="White or Caucasian", 1, 0),
           Race_Other = ifelse(race %in% c("Native Hawaiian or Other Pacific Islander", "Other", "Unknown"), 1, 0))

#m4 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude, covariates = ~ Black + Asian + White + Race_Other)

m4 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
          covariates = ~ as.factor(race))

summary.msm(m4)
```

## Ethnicity

```{r Ethnicity}
df <- df %>% 
    mutate(Hispanic = ifelse(ethnicity=="Hispanic or Latino", 1, 0),
           NotHispanic = ifelse(ethnicity=="Not Hispanic or Latino", 1, 0),
           Ethnicity_Unk = ifelse(ethnicity=="Unknown or Caucasian", 1, 0))

#m5 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude, covariates = ~ Hispanic + NotHispanic + Ethnicity_Unk)

m5 <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
          covariates = ~ as.factor(ethnicity))

summary.msm(m5)
```

## Full model: Race, ethnicity, sex, age, BMI

```{r Full model: Age, BMI, sex, race, ethnicity}
m.full <- msm(state ~ DaysSinceEntry, subject=ID, data=df, qmatrix=Q.crude,
              covariates = ~  age.cen + BMI.cen + as.factor(sex) + as.factor(race) + as.factor(ethnicity))

summary.msm(m.full)

# Transition probability matrix 
pmatrix.msm(m.full)

# Save as R object
# saveRDS(m.full, "/Users/sreynolds2/Documents/GitHub/MS-Markov_Covid_Staging/MSM_full.rds")
```

### Plot of observed and expected prevalences

```{r Full model: Plot of obs and exp prevalences}
plot.prevalence.msm(m.full)
```

# Compare full and baseline model: AIC and log likelihood

```{r Compare full and baseline model}
# AIC
AIC(m0, m.full)

# Log likelihood
logLik(m.full) - logLik(m0)
```

# Other functions from msm package to review

```{r Other functions from msm package, eval=F}
# Rough estimate of goodness of fit of model
    # Estimates observed number of individuals at each state at a series of times
    # Compares these with predicted numbers 
prevalence.msm()

# Plot of observed and expected prevalences 
plot.prevalence.msm(m1b)

# Pearson-type goodness-of-fit test for model
pearson.msm()

# Summarize observed and expected prevalences at each state for each time
    # Also prints hazard ratios for models with covariate effects 
summary.msm(m1a)

# List of tables containing odds ratio estimates, one table per covariate
odds.msm(m1a)
```
